---
title: "P8122hw2_zl2746"
author: "Ze Li"
output: pdf_document
---

```{r library}
library(ggplot2)
library(remotes)
library(permute)
# install_version("ri", "0.9")
library(ri)
```

```{r load data}
light=read.csv("light.csv")
head(light)
summary(light)
```

### part 1: Plot the outcome by treatment group.

```{r Q1}
table(light$Light)
ggplot(light, aes(x = Light, y = BMGain, fill = Light)) +
  geom_boxplot() +
  labs(title = "Body Mass Gain by Light",
       x = "Light",
       y = "Body Mass Gain") +
  theme_minimal()

# Outcome = BMGain & treatment = light
light_LL <- subset(light, Light == 'LL')
ggplot(light_LL, aes(x = BMGain)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Body Mass Gain",
       x = "Body Mass Gain",
       y = "Frequency") +
  theme_minimal()

light_DM <- subset(light, Light == 'DM')
ggplot(light_DM, aes(x = BMGain)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Body Mass Gain",
       x = "Body Mass Gain",
       y = "Frequency") +
  theme_minimal()

light_LD <- subset(light, Light == 'LD')
ggplot(light_LD, aes(x = BMGain)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Body Mass Gain",
       x = "Body Mass Gain",
       y = "Frequency") +
  theme_minimal()
```

### part 2: Subset the data to only consider control = dark & treat = bright.

```{r Q2}
light_LLLD <- subset(light, Light %in% c('LL','LD'))
ggplot(light_LLLD, aes(x = Light, y = BMGain, fill = Light)) +
  geom_boxplot(color = "black") +
  scale_fill_manual(values = c("LD" = "lightcoral", "LL" = "lightblue")) +
  labs(title = "Body Mass Gain by Treatment and Control Group",
       x = "Group (Control: LD, Treatment: LL)",
       y = "Body Mass Gain") +
  theme_minimal()
```

### part 3: Set up the data such that everything you will need has generic names

What quantities will you need to evaluate the causal effect of light at night on weight gain?

Treatment Indicator (A): A binary variable indicating the treatment group (1 for bright light (LL) and 0 for dark light (LD)).

Outcome Variable (Y): The variable measuring the outcome of interest, which in this case is weight gain.

Treatment Effect (T_stat): Compute the difference in means between the treatment (LL) and control groups (LD).

Randomization Inference (Abold): To assess the significance of the observed treatment effect, generate a distribution of treatment effects under the null hypothesis of no causal effect. This involves generating permutations of the treatment assignment.

Distribution of Treatment Effects (rdist): For each permutation of the treatment assignment, calculate the treatment effect.

p-value: Determine the p-value by comparing the observed treatment effect to the distribution of permuted treatment effects.

```{r Q3}
table(light$Light)
# Treatment Indicator (A)
A <- light_LLLD$Light
A <- ifelse(A == "LL", 1, ifelse(A == "LD", 0, A))|>
  as.numeric()
A
# Outcome Variable (Y)
Y <- light_LLLD$BMGain
Y
# Treatment Effect (T_stat)
T_stat <- mean(Y[A == 1]) - mean(Y[A == 0])
T_stat
# Randomization Inference (Abold)
Abold <- genperms(A,maxiter = 12870)
Abold <- genperms(A)
Abold[, 1:6]
# Distribution of Treatment Effects (rdist)
rdist <- rep(NA, times = ncol(Abold))
for (i in 1:ncol(Abold)) {
  A_tilde <- Abold[, i]
  rdist[i] <- mean(Y[A_tilde == 1]) - mean(Y[A_tilde == 0])
}
rdist
hist(rdist)
# p-value
pval <- mean(rdist >= T_stat)
pval
quant <- quantile(rdist,probs = 1-pval)
hist(rdist)
abline(v = quant,col="red")
```

### part 4: the statistic to be the difference in means between the two treatment groups.

```{r Q4}
# Treatment Effect (T_stat)
T_stat <- mean(Y[A == 1]) - mean(Y[A == 0])
T_stat
```

### part 5: How many different possibilities are there for A? Enumerate all of these possibilities in a matrix.

```{r Q5}
n <- length(A)
k <- sum(A) 
A_matrix <- t(combn(n, k, function(x) {
  as.integer(1:n %in% x)
}))

num_pro <- choose(n, k)
num_pro
```

### part 6: State the sharp null hypothesis of no difference. Calculate the test statistic under one of these possibilities for A (the first one), under the sharp null hypothesis.

The sharp null hypothesis states that there is no causal effect of the treatment (light at night) on the outcome (weight gain). 

H0: miu_LL = miu_LD

where miu_LL is the mean weight gain in the treatment group (LL)
and miu_LD is the mean weight gain in the control group (LD).

```{r Q6}
A_first <- A_matrix[1, ]
A_first
T_stat_sharpnull <- mean(Y[A_first == 1]) - mean(Y[A_first == 0])
T_stat_sharpnull
```

### part 7: Generate the exact randomization distribution for T, under the sharp null hypothesis of no difference.

```{r Q7}
T_stats_random <- numeric(nrow(A_matrix)) 

for (i in 1:nrow(A_matrix)) {
  A_tilde <- A_matrix[i, ]  # Each possible treatment assignment
  T_stats_random[i] <- mean(Y[A_tilde == 1]) - mean(Y[A_tilde == 0]) }
```


### part 8: Plot this distribution, and mark the observed test statistic.

```{r Q8}
hist(T_stats_random, 
     main="Randomization Distribution of T",
     xlab="Means difference", 
     breaks=10, 
     col="lightblue", 
     border="black")
abline(v = T_stat, col = "red", lwd = 2, lty = 2)
```

### part 9: Calculate the exact p-value, based on this distribution.

```{r Q9}
p_value_sharp <- mean(abs(T_stats_random) >= abs(T_stat))
p_value_sharp
```

### part 10: What do you conclude?

Since the p-value is smaller than 0.05, we can reject the sharp null hypothesis. It suggests strong statistical evidence that light exposure at night has a significant impact on weight gain in mice.